IML Term project paper

### 2022-12-11 paper.rmd
.

In this project we trained a classifier on a data set of atmospheric measurements. The task is to predict whether new particle formation (NPF) happens or not on a given day based on the atmospheric data. We ran several different kinds of models, assessed through accuracy and perplexity calculations which of these models performed the best, and discuss the results of our analyses.

@@

%Mitä teemme tälle, haluammeko viitata tähän visualisointien ja nimien suhteen vai onko ideaalia, että teemme omat visualisointimme, jotka esittelemme tekstissä?
% (Sara) Ainakin jos tehdään omat visualisoinnit, mielestäni tämän voi poistaa. Visualisointien kanssa kuitenkin miettisin, että mitä todella halutaan näyttää/kertoa, ja tukeeko kyseinen tai kyseiset visualisoinnit tätä. Eli jos visualisointeja halutaan, niin mitä ja miksi, ja miten ne liittyvät luokitteluun ja partikkeleiden muodostumiseen tai muuten tukevat projektityön päämääriä. 
We also familiarized ourselves with the data through the [smear.avaa.scs.fi](https://smear.avaa.csc.fi/) webpage that offers further visualizations details about the data and the measurement site, and with the variable names and details available at [wiki.helsinki.fi](https://wiki.helsinki.fi/pages/viewpage.action?pageId=243959901).

Let's next visualize some our explanatory variables via boxplots and scatterplots to get a deeper understanding of them. For the sake of simplicity, these visualizations grouped by the phenomena being measured. In order to also prelminarily look at how these variables behave in relation to new particle formation events, we'll also look at said measurements behave in relation to variable "class4". In terms of the boxplots, the boxplots are divided into groups by variable class4, and in terms of the scatterplots, the following colors represent different classes: red represents nonevent days, blue represents days with type Ia NPF events, green represents days with type Ib NPF events and black represents days with type II NPF events:
```{r}
cols<-c()
for(i in 1:464){
  if(npf$class4[i]=="nonevent"){
    cols<-c(cols,"red")
  } else if (npf$class4[i]=="Ia"){
    cols<-c(cols,"blue")
  } else if (npf$class4[i]=="Ib"){
    cols<-c(cols,"green")
  } else {
    cols<-c(cols, "black")
  }
}
par(mfrow=c(2,4))
for(i in 2:9){
plot(1:464,npf[,i],col=cols,main = colnames(npf)[i])
}
for(i in 2:9){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(1,2))
for(i in 10:11){
plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 10:11){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(3,4))
for(i in 12:23){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 12:23){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(1,2))
for(i in 24:25){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 24:25){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(3,4))
for(i in 26:37){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 26:37){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(3,4))
for(i in 38:49){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 38:49){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(2,5))
for(i in 50:59){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 50:59){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(2,4))
for(i in 60:67){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 60:67){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(3,4))
for(i in 68:79){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 68:79){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(3,2))
for(i in 80:85){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 80:85){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(2,5))
for(i in 86:95){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 86:95){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(2,2))
for(i in 96:99){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 96:99){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(1,2))
for(i in 100:101){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 100:101){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
```

@@

@@

## Correlation between parameters

@@

As expected, the correlation plot shows that many of the variables describing the same phenomenon at different heights are correlated, as are variables related to radiation. For many classifiers, highly correlated parameters are problematic, and for this reason we would like to drop some of the most highly correlated columns from the dataset.

@@

The correlation plot for the remaining parameter shows us that the remaining hihgly correlated parameters are all radiation-related.

>>>>

In this project we trained a classifier on a data set of atmospheric measurements. The task is to predict whether new particle formation (NPF) happens or not on a given day based on the atmospheric data. We ran several different kinds of models, assessed through accuracy and perplexity calculations which of these models performed the best, and in this report we discuss the process and the results of our analyses.

@@

Let's next visualize some our explanatory variables via boxplots and scatterplots to get a deeper understanding of them. For the sake of simplicity, these visualizations grouped by the phenomena being measured. In order to also prelminarily look at how these variables behave in relation to new particle formation events, we'll also look at said measurements behave in relation to variable "class4". In terms of the boxplots, the boxplots are divided into groups by variable class4, and in terms of the scatterplots, the following colors represent different classes: red represents nonevent days, blue represents days with type Ia NPF events, green represents days with type Ib NPF events and black represents days with type II NPF events:

```{r}
cols<-c()
for(i in 1:464){
  if(npf$class4[i]=="nonevent"){
    cols<-c(cols,"red")
  } else if (npf$class4[i]=="Ia"){
    cols<-c(cols,"blue")
  } else if (npf$class4[i]=="Ib"){
    cols<-c(cols,"green")
  } else {
    cols<-c(cols, "black")
  }
}
par(mfrow=c(2,4))
for(i in 2:9){
plot(1:464,npf[,i],col=cols,main = colnames(npf)[i])
}
for(i in 2:9){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}

```
(Other visualizations omitted from the preliminary report.)
```{r, eval=FALSE}
cols<-c()
for(i in 1:464){
  if(npf$class4[i]=="nonevent"){
    cols<-c(cols,"red")
  } else if (npf$class4[i]=="Ia"){
    cols<-c(cols,"blue")
  } else if (npf$class4[i]=="Ib"){
    cols<-c(cols,"green")
  } else {
    cols<-c(cols, "black")
  }
}
par(mfrow=c(2,4))
for(i in 2:9){
plot(1:464,npf[,i],col=cols,main = colnames(npf)[i])
}
for(i in 2:9){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(1,2))
for(i in 10:11){
plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 10:11){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(3,4))
for(i in 12:23){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 12:23){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(1,2))
for(i in 24:25){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 24:25){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(3,4))
for(i in 26:37){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 26:37){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(3,4))
for(i in 38:49){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 38:49){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(2,5))
for(i in 50:59){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 50:59){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(2,4))
for(i in 60:67){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 60:67){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(3,4))
for(i in 68:79){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 68:79){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(3,2))
for(i in 80:85){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 80:85){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(2,5))
for(i in 86:95){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 86:95){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(2,2))
for(i in 96:99){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 96:99){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(1,2))
for(i in 100:101){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 100:101){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
```

@@

## Correlation

### Correlation between predictors and class

@@

### Correlation between parameters

@@

As expected, the correlation plot shows that many of the variables describing the same phenomenon at different heights are correlated, as are variables related to radiation. For many classifiers, highly correlated parameters are problematic, and for this reason we would like to drop as many of these columns from the dataset as possible.

@@

The correlation plot for the remaining parameter shows us that the remaining highly correlated parameters are all radiation-related.
